Performance analysis of high-resolution \\ ice sheet simulations

### 2022-07-04 paper/perfmod.tex
this version posted to arxiv ... email friends once processed
posted at arxiv (https://arxiv.org/abs/2206.14352) and submitted to J Glaciol
I submitted it again in the review format
correct errors (alpha=1 not 2 for sparsity-exploiting direct LA; beta=0.8 not beta=1.8 for Bueler 2016 effort)

%\documentclass[twocolumn,letterpaper]{igs}
\documentclass[twocolumn]{igs}  % A4 paper needs no option
%\documentclass[review,letterpaper]{igs}

@@

%\keywords{}

@@

This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.  Our simplified performance analysis can expose the most important considerations and trade-offs.  While the author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, these beliefs should be inspected quantitatively to the extent possible.

@@

For each explicit time step of a model using a membrane-stress-resolving balance the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 2$ if sparsity is exploited \citep{Bueler2021}.

@@

Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.  At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula; an SIA velocity solution requires optimal $O(m)$ flops.

When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, strongly depending on solver design.  Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling.  Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.  For example, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good parallel scaling, something we do not address.

@@

Now, how many flops are needed to simulate one model year?  Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ time steps per model year.  We may write stability restrictions as required numbers of time-steps per model year.  That is, $\Delta t^{-1}$ is bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.  Recalling scaling \eqref{eq:paramrelation} of $\Delta x$ with $m$, in the explicit SIA and pessimistic-Stokes cases we have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}
The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

@@

The only implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.  For simplified dome geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=1.8$.  (An earlier non-Newton implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.  Practical steady-state ($\Delta t=\infty$) solutions at ice sheet scale are also demonstrated by \cite{JouvetBueler2012} and \cite{Bueler2016}.

The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are also solved implicitly.  \cite{JouvetGraeser2013} use a multigrid method, but their published, simplified-geometry results do not constrain $\beta$.  Overall time-stepping in these hybrids is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects the thickness, so the SIA-portion solvers are apparently only tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$ with $U$ scaling according to the sliding.
% equation (20) in JouvetGraeser2013
% equation (A21) in BrinkerhoffJohnson2015

For implicit Stokes time-stepping, a prospective, coupled, and free-boundary velocity and geometry-update solve is assumed to be, in the absense of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.  One might also suppose $\gamma\ge \beta$, but in any case there are no implemented cases to measure.  These comments complete Table \ref{tab:performancemodel}, in which all estimates involve a scheme-dependent constant, something which is especially large for the implicit schemes.

@@

From Table \ref{tab:performancemodel} we first observe a known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that effort scales as $O(\Delta x^{-4})$.  Recall that this follows because $\Delta t < O(\Delta x^2)$, and because the expense of one geometry-update operation is $O(m) = O(\Delta x^{-2})$.  Spatial mesh refinement by a factor of two therefore imposes an impressive 16-times increase in effort.

The \cite{Bueler2016} implicit SIA solver is really no better, however.  Although long time steps can be taken by this implicit solver, the cost of $\beta=1.8$ scaling is $O(\Delta x^{-5.6})$ because $2+2\beta=5.6$.  However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably superior performance.

Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of solvers completely dominates.  In particular, an $\alpha=2$ explicit Stokes model, e.g.~one using direct linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-8})$ under a pessimistic stability condition.  Optimistic stability yields still-horrific $O(\Delta x^{-7})$.  These simple observations emphasize the key role of multigrid-based stress balance solvers.  That is, even retaining explicit time-stepping, nearly-optimal ($\alpha \approx 0$) scaling of computation effort in Stokes and higher-order solvers will be necessary for routine application on high-resolution meshes.

On the other hand, suppose resolution ($\Delta x$) is fixed.  The Table also shows why algorithmic scaling remains important in the large-domain $L\to\infty$ limit when applying Stokes or higher-order dynamics.  An $\alpha=2$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.  By contrast, the computational work of a nearly-optimal solver will be proportional to ice sheet area $L^2$.

The promise of nearly-optimal solvers is profoundly revealed when implicit geometry updates are considered.  Future methods which simultanously satisfy the mass and momentum equations at each time step, and which do work essentially proportional to the number of degrees of freedom, have the greatest promise.  A $\gamma \approx 0$ implicit Stokes method would be hugely more capable for many tasks.  Specifically, and more achievably, a $\gamma \le 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method, presumably based on multigrid solution of the free-boundary problem for the coupled mass and momentum equations, is an appropriate goal for decades of research on numerical ice sheet models.  The computation cost would scale at $O(\Delta x^{-3})$, better than explicit SIA models, and the method would only update ice geometry when determined by the scientific need to resolve climate coupling, while avoiding shallow approximations.  The same goal makes sense for all membrane-stress-resolving solvers.  No apparent technical progress has yet been made on such an implicit Stokes design, and so these aspirations are decidedly long-term.  However, the above discussion suggests why measured values for $\alpha,\beta,\gamma$, or equivalent algorithmic scaling measures, are important performance metrics to report when describing new ice sheet solvers.

>>>>

\documentclass[twocolumn,letterpaper]{igs}
%\documentclass[twocolumn]{igs}  % A4 paper needs no option
%\documentclass[review,letterpaper]{igs}

@@

%\keywords{glacier flow,glacier modelling,ice-sheet modelling}
% suggested reviewers:  Dan Goldberg, William Lipscomb, Toby Isaac, Josefin Ahlkrona, Alex Jarosch, Doug Brinkerhoff

@@

This is the context in which the current note relates time-stepping and stress-balance choices to computational effort.  Our simplified performance analysis exposes the most important considerations and trade-offs.  While the author expects that implicit time-stepping ice sheet models will eventually be the fastest, and that advanced solver techniques like multigrid \citep{Briggsetal2000} will be necessary, these beliefs should be assessed quantitatively to the extent possible.

@@

For each explicit time step of a model using a membrane-stress-resolving balance the computational cost of a velocity (or velocity/pressure) solution of the stress balance equations is determined by solver design.  We suppose that one such solution requires $O(m^{1+\alpha})$ floating point operations (flops),  with the power $\alpha\ge 0$ depending on the solver implementation.  For example, a Stokes solver using direct linear algebra for each Newton step might yield $\alpha \approx 1$ if sparsity is exploited or $\alpha \approx 2$ if not \citep{Bueler2021}.

@@

Note that $\alpha=0$ describes an \emph{optimal} solver in the language of algorithmic scaling or solver complexity \citep{Bueler2021}.  At the easiest end of ice sheet modeling, the SIA velocity computation is a trivialization of the Stokes problem in which velocity is computed by a pointwise formula; an SIA velocity solution therefore requires optimal $O(m)$ flops.

When analyzing solver scaling in our simplified form, one must be aware that the constant in $O(m^{1+\alpha})$ can be very large, strongly depending on solver design.  Furthermore, many considerations are suppressed in any flops-based analysis of algorithmic scaling.  Actual run time is also determined by memory latency, memory bandwidth, and process/thread/GPU parallelism, among other factors.  For example, the \cite{BrownSmithAhmadia2013,Fischleretal2022,IsaacStadlerGhattas2015,Lengetal2012,Tuminaroetal2016} results, among others, all show good parallel scaling, something we do not address here.

@@

Now, how many flops are needed to simulate one model year?  Suppose a numerical model takes time steps of $\Delta t$ model years, equivalently $\Delta t^{-1}$ steps per model year.  We may write stability restrictions as required numbers of steps per model year.  That is, $\Delta t^{-1}$ is bounded below by a function of the horizontal resolution $\Delta x$ or the degrees of freedom $m$.  Recalling scaling \eqref{eq:paramrelation} of $\Delta x$ with $m$, in the explicit SIA and pessimistic-Stokes cases we have
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{D}{\Delta x^2}} = \oo{\frac{D m}{L^2}}. \label{eq:explicitsiarequired}
\end{equation}
The explicit, optimistic-Stokes estimate becomes
\begin{equation}
\frac{1}{\Delta t} > \oo{\frac{U}{\Delta x}} = \oo{\frac{U m^{1/2}}{L}}. \label{eq:explicitoptstokesrequired}
\end{equation}
The number of time steps per model year is then multiplied by the per-step computational cost, namely $O(m^{1+\alpha})$, to give a work estimate for each model year in a simulation.  The results so far are shown in the ``explicit'' rows of Table \ref{tab:performancemodel}.

@@

The only implemented, unconditionally-stable, fully-implicit geometric-update solvers use the SIA stress balance.  For simplified dome geometry the scheme in \cite{Bueler2016}, based on Newton steps solved via direct linear algebra and single-grid ice margin determination, shows $\beta=0.8$.  (An earlier non-Newton implementation by \cite{JouvetBueler2012} scales worse.)  Convergence is robust for time steps of years to centuries on kilometer-scale grids for the Greenland ice sheet, using realistic, thus irregular, bed topographies.  Practical steady-state ($\Delta t=\infty$) solutions at ice sheet scale are also demonstrated by \cite{JouvetBueler2012} and \cite{Bueler2016}.

The SIA portions of hybrid time-stepping schemes by \cite{JouvetGraeser2013} and \cite{BrinkerhoffJohnson2015} are also solved implicitly.  \cite{JouvetGraeser2013} use a multigrid method, but their published, simplified-geometry results do not constrain $\beta$.  Overall time-stepping in these hybrids is only semi-implicit because the membrane-stress-resolving portion of the velocity explicitly advects the thickness, so the SIA-portion solvers are apparently only tested for time steps satisfying an advective condition $\Delta t < O(U^{-1}\Delta x)$ for $U$ which scales according to the sliding.
% equation (20) in JouvetGraeser2013
% equation (A21) in BrinkerhoffJohnson2015

For implicit Stokes time-stepping, a prospective, coupled, and free-boundary velocity and geometry-update solve is assumed to be, in the absence of constraining research, $O(m^{1+\gamma})$ for some $\gamma \ge \alpha$ to be determined.  One might also suppose $\gamma\ge \beta$, but in any case there are no implemented cases to measure.  These comments complete Table \ref{tab:performancemodel}, in which all estimates involve a scheme-dependent constant, something which is especially large for the implicit schemes.

@@

From Table \ref{tab:performancemodel} we first observe a known property of explicit time-stepping for 2D (map-plane) diffusion equations such as SIA equation \eqref{eq:siamasscontinuity}, namely that effort, here flops per model year, scales as $O(\Delta x^{-4})$.  Recall that this follows because $\Delta t < O(\Delta x^2)$, and because the expense of one geometry-update operation is $O(m) = O(\Delta x^{-2})$.  Spatial mesh refinement by a factor of two therefore imposes an impressive 16-times increase in effort.

The \cite{Bueler2016} implicit SIA solver is not enough better, however.  Although long time steps can be taken by this implicit solver, the $\beta=0.8$ scaling gives $O(\Delta x^{-3.6})$ effort because $2+2\beta=3.6$.  However, an improvement to $\beta < 0.5$, presumably by application of a multigrid method, would transform such an implicit solver into a tool with notably superior performance compared to an explicit, $O(\Delta x^{-4})$ SIA scheme.

Now bypassing the over-simplified SIA stress balance, we see from Table \ref{tab:performancemodel} how the algorithmic scaling of solvers completely dominates.  In particular, an $\alpha=1$ explicit Stokes model, e.g.~one using direct, sparsity-exploiting linear algebra on each Newton step system, will do work which scales at the horrific rate $O(\Delta x^{-6})$ under a pessimistic stability condition.  Optimistic stability yields still-bad $O(\Delta x^{-5})$.  These simple observations emphasize the key role of multigrid-based stress balance solvers.  That is, even retaining explicit time-stepping, nearly-optimal ($\alpha \approx 0$) scaling of computation effort in Stokes and higher-order solvers will be necessary for routine application on high-resolution meshes.

On the other hand, suppose resolution ($\Delta x$) is fixed.  The Table also shows why algorithmic scaling remains important in the large-domain $L\to\infty$ limit when applying Stokes or higher-order dynamics.  An $\alpha=1$ method which might suffice for a smaller $L=100$ km ice cap will struggle for a $L=1000$ km ice sheet because the effort scales as the fourth ($2+2\alpha=4$) power of $L$.  By contrast, the computational work of a nearly-optimal solver will be proportional to ice sheet area $L^2$.

The promise of nearly-optimal solvers is profoundly revealed when implicit geometry updates are considered.  Future methods which simultanously satisfy the mass and momentum equations at each time step, and which do work essentially proportional to the number of degrees of freedom, have the greatest promise.  A $\gamma \approx 0$ implicit Stokes method would be hugely more capable for many tasks.  Specifically, and more achievably, a $\gamma < 0.5$ implicit, essentially unconditionally-stable, Stokes time-stepping method, presumably based on multigrid solution of the free-boundary problem for the coupled mass and momentum equations, is an appropriate goal for coming decades of research on numerical ice sheet models.  The computation cost would scale at $O(\Delta x^{-3})$, better than explicit SIA models, and the method would only update ice geometry when determined by the scientific need to resolve climate coupling, while avoiding shallow approximations.  The same goal makes sense for all membrane-stress-resolving solvers.  No apparent technical progress has yet been made on such an implicit Stokes design, and so these aspirations are decidedly long-term.  However, the above discussion suggests why measured values for $\alpha,\beta,\gamma$, or equivalent algorithmic scaling measures, are important performance metrics to report when describing new ice sheet solvers.

\subsubsection{Note added 4 July 2022.}  Errors in the first posted version (\url{https://arxiv.org/abs/2206.14352}) are corrected here.
